\pkg{casebase}: An Alternative Framework For Survival Analysis

### 2017-09-18 paper/casebase_jss.Rmd

The data analysis step was separated into three parts: 1) case-base sampling; 2) estimation of the smooth hazard function; and 3) calculation of the risk function. By separating the sampling and estimation functions, we allowed the possibility of users implementing more complex sampling scheme, as described in Saarela.

The sampling scheme selected for \code{sampleCaseBase} was described in Hanley and Miettinen: we first sample along the ``person'' axis, proportional to each individual's total follow-up time, and then we sample a moment uniformly over their follow-up time. This sampling scheme is equivalent to the following picture: imagine representing the total follow-up time of all individuals in the study along a single dimension, where the follow-up time of the next individual would start exactly when the follow-up time of the previous individual ends. Then the base series could be sampled uniformly from this one-dimensional representation of the overall follow-up time. In any case, the output is a dataset of the same class as the input, where each row corresponds to a person-moment. The covariate profile for each such person-moment is retained, and an offset term is added to the dataset. This output could then be used to fit a smooth hazard function, or for visualization of the base series.

>>>>

The data analysis step was separated into three parts: 1) case-base sampling; 2) estimation of the smooth hazard function; and 3) calculation of the risk function. By separating the sampling and estimation functions, we allowed the possibility of users implementing more complex sampling scheme, as described in Saarela [-@saarela2016case].

The sampling scheme selected for \code{sampleCaseBase} was described in Hanley and Miettinen [-@hanley2009fitting]: we first sample along the ``person'' axis, proportional to each individual's total follow-up time, and then we sample a moment uniformly over their follow-up time. This sampling scheme is equivalent to the following picture: imagine representing the total follow-up time of all individuals in the study along a single dimension, where the follow-up time of the next individual would start exactly when the follow-up time of the previous individual ends. Then the base series could be sampled uniformly from this one-dimensional representation of the overall follow-up time. In any case, the output is a dataset of the same class as the input, where each row corresponds to a person-moment. The covariate profile for each such person-moment is retained, and an offset term is added to the dataset. This output could then be used to fit a smooth hazard function, or for visualization of the base series.

### 2017-09-18 paper/casebase_jss.tex

\begin{enumerate}
\def\labelenumi{\arabic{enumi}.}
\tightlist
\item
  Population-tim plots
\item
  Sampling
\item
  Fitting
\item
  Absolute Risks
\end{enumerate}

\section{Population-time plots}\label{population-time-plots}

@@

\begin{tabular}{l|r|l}
\hline
Covariates & HR & 95\% CI\\
\hline
Sex & 0.71 & (0.41, 1.24)\\
\hline
Disease & 0.52 & (0.28, 0.93)\\
\hline
Phase (CR2 vs. CR1) & 1.17 & (0.47, 2.93)\\
\hline
Phase (CR3 vs. CR1) & 1.78 & (0.46, 6.91)\\
\hline
Phase (Relapse vs. CR1) & 4.45 & (2.07, 9.57)\\
\hline
Source & 1.64 & (0.54, 5)\\
\hline
Age & 0.99 & (0.97, 1.02)\\
\hline
\end{tabular}

@@

\section{Discussion}\label{discussion}

\bibliography{casebase_references.bib}

>>>>

The functions in the casebase package can be divided into two
categories: 1) data visualization, in the form of population-time plots;
and 2) data analysis.

We explicitely aimed at being compatible with both \code{data.frame}s
and \code{data.table}s. This is evident in some of the coding choices we
made, and it is also reflected in our testing units.

\subsection{Population-time plots}\label{population-time-plots}

\subsection{Data analysis}\label{data-analysis}

The data analysis step was separated into three parts: 1) case-base
sampling; 2) estimation of the smooth hazard function; and 3)
calculation of the risk function. By separating the sampling and
estimation functions, we allowed the possibility of users implementing
more complex sampling scheme, as described in Saarela
\citeyearpar{saarela2016case}.

The sampling scheme selected for \code{sampleCaseBase} was described in
Hanley and Miettinen \citeyearpar{hanley2009fitting}: we first sample
along the ``person'' axis, proportional to each individual's total
follow-up time, and then we sample a moment uniformly over their
follow-up time. This sampling scheme is equivalent to the following
picture: imagine representing the total follow-up time of all
individuals in the study along a single dimension, where the follow-up
time of the next individual would start exactly when the follow-up time
of the previous individual ends. Then the base series could be sampled
uniformly from this one-dimensional representation of the overall
follow-up time. In any case, the output is a dataset of the same class
as the input, where each row corresponds to a person-moment. The
covariate profile for each such person-moment is retained, and an offset
term is added to the dataset. This output could then be used to fit a
smooth hazard function, or for visualization of the base series.

The fitting function \code{fitSmoothHazard} starts by looking at the
class of the dataset: if it was generated from \code{sampleCaseBase}, it
automatically inherited the class \code{cbData}. If the dataset supplied
to \code{fitSmoothHazard} does not inherit from \code{cbData}, then the
fitting function starts by calling \code{sampleCaseBase} to generate the
base series. In other words, the occasional user can bypass
\code{sampleCaseBase} altogether and only worry about the fitting
function \code{fitSmoothHazard}.

The fitting function retains the familiar formula interface of
\code{glm}. The left-hand side of the formula should be the name of the
column corresponding to the event type. The right-hand side can be any
combination of the covariates, along with an explicit functional form
for the time variable. Note that non-proportional hazard models can be
achieved at this stage by adding an interaction term involving time. The
offset term does not need to be specified by the user, as it is
automatically added to the formula.

Finally, the hazard function is fitted to the data using the function
\code{glm}, unless there is more than one type event (i.e.~in a
competing-risk analysis), in which case it uses the multinomical
regression capabilities of the \pkg{VGAM} package. This package was
selected for its ability to fit multinomial regression models with an
offset.

Once a model-fit object has been returned by \code{fitSmoothHazard}, all
the familiar summary and diagnostic functions are available:
\code{print}, \code{summary}, \code{predict}, \code{plot}, etc. Our
package provides one more functionality: it computes risk functions from
the model fit. For the case of a single event, it uses the familiar
identity \[S(t) = \exp\left(-\int_0^t h(u;X) du\right).\] The integral
is computed using either the \code{stats::integrate} function or
Monte-Carlo integration. For the case of a competing-event analysis, the
event-specific risk is computed using a nested double integral; in this
setting, Monte-Carlo integration is faster and more flexible than
\code{stats::integrate}. This is due to the fact that the computation
involves a double loop over the selected time points; Monte-Carlo
integration performs this step by using the vectorized \code{rowSums}
and \code{colSums} functions.

To decide between a single-event and a competing-event analysis, we
created \code{absoluteRisk} as an \code{S3} generic, with methods for
both \code{glm} and \code{CompRisk} objects (the latter inherits from
\code{vglm} as well). The method dispatch system of \proglang{R} then
takes care of matching the correct input to the correct methodology,
without the user's intervention.

@@

\begin{tabular}{l|r|l}
\hline
Covariates & HR & 95\% CI\\
\hline
Sex & 0.81 & (0.47, 1.41)\\
\hline
Disease & 0.57 & (0.31, 1.03)\\
\hline
Phase (CR2 vs. CR1) & 1.21 & (0.48, 3.01)\\
\hline
Phase (CR3 vs. CR1) & 1.75 & (0.45, 6.8)\\
\hline
Phase (Relapse vs. CR1) & 4.23 & (1.97, 9.09)\\
\hline
Source & 1.60 & (0.53, 4.88)\\
\hline
Age & 1.00 & (0.97, 1.02)\\
\hline
\end{tabular}

@@

\renewcommand\refname{Discussion}
\bibliography{casebase_references.bib}
